### 2020-07-10 case-studies.tex
finished first pass through sec 3
removing todos

\section{Case Studies}\label{sec:case-studies}
Let us begin by considering livelits from the client programmer's perspective by way of 
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading} 
and an image transformation case study in Sec.~\ref{sec:image-transformation}. 
We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}. 
These case studies and examples have been implemented\todo{make this not a lie}{}
in Hazel, a browser-based live programming environment for a pure typed functional language in the
ML family. We assume basic familiarity with ML.\todo{cite Elm tutorial for readers who are not familiar?}{}

\subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
\begin{figure}
\begin{lstlisting}
let weights = [0.1, 0.1, 0.1, 0.1, 0.25, 0.35] in 
let grades = $dataframe in 
            A1          A2          A3          A4       Midterm      Final  +
  Cyrus
  Nick
  David
    + 
let averages = compute_averages(grades, weights) in 
let cutoffs = $cutoffs(averages) in

  |---------------------------------------------------|

format_for_university(assign_grades(averages, cutoffs))
\end{lstlisting}
\caption{Grading Example}
\label{fig:grading}
\end{figure}

@@

Perhaps the most common end-user application in this domain is the spreadsheet.
A spreadsheet allows the instructor to record student grades using a natural tabular interface,
visualize this data in one of a finite number of styles, and perform basic computations on this data,
with results updated live, i.e. automatically after each cell has been edited. 
However, these affordances are limited. It is difficult to package up common operations and workflows 
into reusable libraries, interact with the data using domain-specific visualizations, 
and perform operations that have not been anticipated by the designers of the spreadsheet system 
(e.g. operations that prepare the data in an idiosyncratic format suitable for automated submission 
to the university's grading system). 
General-purpose programming languages equipped with suitable libraries for working with numeric data
are flexible enough to handle these scenarios, but users of these languages 
sacrifice the ability to receive live feedback and directly manipulate the data and visualizations thereof.

Livelits are able to address this tension. 
Fig.~\ref{fig:grading} shows a Hazel program where the user has:\todo{label each of these 1-5 in figure?}{}
\begin{enumerate}
\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expression in each cell, 
\item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses, 
\item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution 
(for the sake of space, we plot dots rather than a full histogram)\todo{should we do a histogram?}{}, 
\item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades}, 
\item formatted the final grade assignments in a manner suitable for submission to the university system by calling another shared library function, \li{format_for_university}.
\end{enumerate}

\subsubsection{Livelit Expansion} 
Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
For example, the expansion of Fig.~\ref{fig:grading} is:\todo{change code once figure is final}{}

\begin{lstlisting}
let grades = Dataframe (
  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
  [
    ("Cyrus", [0.0, 0.0, 0.0, 0.0, 0.0]),
    ("Nick", [0.0, 0.0, 0.0, 0.0, 0.0]),
    ("David", [0.0, 0.0, 0.0, 0.0, 0.0])
  ]) in 
let averages = compute_averages(grades) in 
let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in 
format_for_university(assign_grades(averages, cutoffs))
\end{lstlisting}

The programmer can inspect this expansion in Hazel. 
Ideally, however, it would not be necessary to inspect the expansion 
(or the livelit implementation, which specifies the expansion logic and is discussed in Sec.~\ref{sec:livelit-definitions}) 
to reason about types and binding. 
After all, one does not need to look inside 
function bodies to reason about types and binding.
Instead, in the words of Reynolds\todo{cite}{}, 
``type structure is a syntactic discipline for maintaining levels of abstraction''.
The livelits mechanism supports reasoning abstractly about types and binding by 
several means.

To support abstract reasoning about the type of the expansion, 
livelit definitions declare an \emph{expansion type}.
The declarations of the two livelits in Fig.~\ref{fig:grading}, 
eliding their implementations, are:
\begin{lstlisting}[numbers=none]
livelit $dataframe at Dataframe { ... }
livelit $cutoffs(avgs: List(Float)) 
  at (.A Float, .B Float, .C Float, .D Float) { ... }
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe} 
(a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
and the expansion type of \li{\$cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
Hazel summarizes the typing information from the livelit definition for the programmer when the cursor is on the livelit's name, 
just as it displays a function's type when its name is under the cursor (not shown).\todo{could show if Hannah et al are finished}{}

@@

\paragraph{Parameters}\label{sec:parameterization} Parameters appear immediately after the name of the applied livelit. 
For example, \li{averages} is provided as a parameter to \li{\$cutoffs} in Fig.~\ref{fig:grading}.
The livelit declares a fixed number of parameters and specifies their types in its declaration. 
For example, the declaration of \li{\$cutoffs} shown above specifies one parameter of type \li{List(Float)}
(the grades data to be plotted, see below).

@@

To avoid this situation, the livelit expansion mechanism 
maintains parameter and splice hygiene: parameters and splices are included in the expansion 
in a capture-avoiding manner.
Consequently, clients can reason abstractly about binding in splices: variables in splices
always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
This is implemented by alpha-renaming bindings internal to the expansion as necessary.
(We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)

\subsubsection{Live Evaluation}\label{sec:live-evaluation}
Livelits can request the result of evaluating a splice or a parameter. 
For example, the \li{\$dataframe} livelit uses this facility to display
the evaluation result for each cell when the cursor is not on it, like a spreadsheet.
The \li{\$cutoffs} livelit uses this facility to plot the grades list, provided
as a parameter, on the number line.

@@

Similarly, the closure for the \li{\$cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for 
the \li{grades} variable. However, this variable's result, along with the 
results for other variables that depend on \li{grades}, such as \li{averages}, ultimately depend on the 
expansion of the \li{\$dataframe} livelit. If we stop after the first phase of closure collection,
where these expansions have not yet been included, 
then no useful result will be available here.
For this reason, there is a second phase of closure collection where any livelit holes that appear 
in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated 
(in this case, for the \li{\$dataframe} livelit) and is used to fill 
the hole before evaluation resumes, following the semantics for resumption developed by \citet{HazelnutLive}.
Livelit expansions do not themselves contain livelits, so no subsequent resumption phases are necessary.

@@

\paragraph{Indeterminate Results}
We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully: 
splices and parameters can be incomplete, i.e. they can contain holes, so 
evaluation does not always result in a value \cite{HazelnutLive}.
In particular, when a hole appears in elimination position, e.g. in function position of a function application,
 the result is instead an indeterminate expression. 
When a livelit requests an evaluation result, it must be able to handle indeterminate results.
For example, if the parameter to \li{\$cutoffs} were a hole or some other indeterminate expression\todo{small screenshot?}{}, 
then it would have degraded functionality 
(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements\todo{implement this}{}). 
We will return to how livelit implementations handle this situation in Sec.~\ref{sec:implementation}.
\todo{make sure we cover this}{}

>>>>

\section{Case Studies}\label{sec:case-studies}
Let us begin by considering livelits from the client programmer's perspective by way of
two domain-specific case studies:
a course grade assignment case study in Sec.~\ref{sec:live-grading}
and an image transformation case study in Sec.~\ref{sec:image-transformation}.
We also briefly mention several other examples in Sec.~\ref{sec:additional-examples}.\todo{do we?}{}
These case studies and examples have been implemented
in Hazel, a browser-based live programming environment for a pure typed functional language in the
ML family. We assume basic familiarity with ML.

\subsection{Case Study: Grading with Livelits}\label{sec:live-grading}
\begin{figure}
    \includegraphics[width=30pc]{grade-cutoff-livelit.png}
\caption{Grading with Livelits}
\label{fig:grading}
\end{figure}

@@

Perhaps the most common end-user application in this domain is the spreadsheet.
A spreadsheet allows the instructor to record student grades using a natural tabular interface,
visualize this data in one of a finite number of plot styles, and perform basic computations on this data,
with results updated live, i.e. automatically after each cell has been edited.
However, these affordances are limited. It is difficult to package up common operations and workflows
into reusable libraries, interact with the data using domain-specific visualizations,
and perform operations that have not been anticipated by the designers of the spreadsheet system
(e.g. operations that prepare the data in an idiosyncratic format suitable for automated submission
to the university's grading system).
General-purpose programming languages equipped with suitable libraries for working with numeric data
are flexible enough to handle these scenarios, but users of these languages
sacrifice the ability to receive live feedback and directly manipulate data and visualizations thereof.

Livelits are able to address this tension.
Fig.~\ref{fig:grading} shows a Hazel program where the user has:
\begin{enumerate}
\item recorded numeric \li{grades} for each student using a livelit, \li{\$dataframe}, that implements a tabular user interface with support for live evaluation of the Hazel expression in each cell,
\item programmatically computed the overall \li{averages} for each student by applying a function, \li{compute_averages}, defined in a library  (not shown) shared between multiple courses,
\item determined reasonable \li{cutoffs} between letter grades by directly manipulating a livelit that offers a domain-specific drag-and-drop interface superimposed on a live visualization of the grade distribution
(for the sake of space, we plot dots rather than a full histogram),
\item programmatically assigned letter grades to each student based on the \li{cutoffs} determined in step (3) by calling another shared library function, \li{assign_grades},
\item formatted the final grade assignments in a manner suitable for submission to the university system by calling another shared library function, \li{format_for_university}.
\end{enumerate}

\subsubsection{Livelit Expansion}\label{sec:livelit-expansion}
Livelit expressions are given dynamic meaning by expansion to core expressions, i.e. expressions that do not contain livelits.
For example, the expansion of Fig.~\ref{fig:grading} is:

\begin{lstlisting}
let grades = Dataframe (
  ["A1", "A2", "A3", "A4", "Midterm", "Final"],
  [
    ("Alice", [24. +. 36. +. 33., 92., 83.5, 87.5, 95., 88.]),
    ("Bob", [61., 64., 98., 89.5, 70., 85.]),
    ("Carol", [75., 81., 73., 71., 82., 79.]),
    (* ... *)
  ]) in
let averages = compute_averages grades weights in
let cutoffs = (.A 90.0, .B 80.0, .C 70.0, .D 60.0) in
format_for_university (assign_grades averages cutoffs)
\end{lstlisting}

The programmer can inspect this expansion in Hazel.
Ideally, however, it would not be necessary to inspect the expansion
(or the livelit implementation, which specifies the expansion logic and is discussed in Sec.~\ref{sec:livelit-definitions})
to reason about types and binding.
After all, one does not need to look inside
function bodies to reason about types and binding.
Instead, in the words of \citet{DBLP:conf/ifip/Reynolds83},
``type structure is a syntactic discipline for maintaining levels of abstraction''.
The livelits mechanism supports reasoning abstractly about types and binding by
several means.

To support abstract reasoning about the type of the expansion,
livelit definitions declare an \emph{expansion type}.
The declarations of the two livelits in Fig.~\ref{fig:grading},
eliding their implementations, are:
\begin{lstlisting}[numbers=none]
livelit $dataframe at Dataframe { ... }
livelit $grade_cutoffs(avgs: List(Float))
  at (.A Float, .B Float, .C Float, .D Float) { ... }
\end{lstlisting}
The expansion type of \li{\$dataframe} is \li{Dataframe}
(a type classifying tabular floating point data together with string row and column keys, consistent with the expansion above)
and the expansion type of \li{\$grade_cutoffs} is a labeled product of grade cutoffs (Hazel notates labels \li{.label}).
Hazel summarizes the typing information from the livelit definition when the cursor is on the livelit's name,
just as it displays a function's type when its name is under the cursor (not shown).

@@

\paragraph{Parameters}\label{sec:parameterization} Parameters appear immediately after the name of the applied livelit.
For example, \li{averages} is provided as a parameter to \li{\$grade_cutoffs} in Fig.~\ref{fig:grading}.
The livelit declares a fixed number of parameters and specifies their types in its declaration.
For example, the declaration of \li{\$grade_cutoffs} shown above specifies one parameter of type \li{List(Float)}
(the grades data to be plotted, see below).

@@

To avoid this situation, the livelit expansion mechanism
maintains parameter and splice hygiene: parameters and splices are included in the expansion
in a capture-avoiding manner.
Consequently, clients can reason abstractly about binding in splices: variables in splices
always refer to bindings visible to the client, rather than bindings that are hidden inside expansions.
% This is implemented by alpha-renaming bindings internal to the expansion as necessary.
% (We discuss potentially relaxed variations of this hygiene discipline in Sec.~\ref{sec:discussion}.)

\subsubsection{Live Evaluation}\label{sec:live-evaluation}
Livelits can request the result of evaluating a splice or a parameter.
For example, the \li{\$dataframe} livelit uses this facility to display
the evaluation result for each cell when the cursor is not on it, like a spreadsheet.
The \li{\$grade_cutoffs} livelit uses this facility to plot the grades list, provided
as a parameter, on the number line.

@@

Similarly, the closure for the \li{\$grade_cutoffs} livelit in Fig.~\ref{fig:grading} includes a result for
the \li{grades} variable. However, this variable's result, along with the
results for other variables that depend on \li{grades}, such as \li{averages}, ultimately depend on the
expansion of the \li{\$dataframe} livelit. If we stop after the first phase of closure collection,
where these expansions have not yet been included,
then no useful result will be available here.
For this reason, there is a second phase of closure collection where any livelit holes that appear
in the collected livelit closures are \emph{resumed}, i.e. the expansion is generated
(in this case, for the \li{\$dataframe} livelit) and is used to fill
the hole before evaluation resumes, following the semantics for resumption developed by \citet{HazelnutLive}.
Livelit expansions do not themselves contain livelits, so no subsequent resumption phases are necessary.

@@

\paragraph{Indeterminate Results}
We used the phrase \emph{evaluation result} rather than \emph{value} above purposefully:
splices and parameters can be incomplete, i.e. they can contain holes, so
evaluation does not always result in a value \cite{HazelnutLive}.
In particular, when a hole appears in elimination position, e.g. in function position of a function application,
 the result is instead an indeterminate expression.
When a livelit requests an evaluation result, it must be able to handle indeterminate results.
For example, if the parameter to \li{\$grade_cutoffs} were a hole or some other indeterminate expression,
then it would have degraded functionality
(in this case, it would display the list elements that are values on the timeline, skipping indeterminate elements).
We will return to how livelit implementations handle this situation in Sec.~\ref{sec:livelit-definitions}.

### 2020-07-10 intro.tex
Restore subfigure labels
removing todos

\subsection{Background}
\begin{figure*}
  \begin{center}
    \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{slider-color-livelits.png}
    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{slider-color-livelits.png}
    \end{subfigure}
  \end{center}
  \caption{
  (a) This figure, adapted from the prior work of \citet{Graphite},
  shows a Graphite palette associated with the \texttt{Color} class.
  The client can specify the RGBA components only using literal numbers.
  (b) A similar Hazel livelit for colors uses live splices for the RGBA components,
  so the client can enter any Hazel expression.
  In this case, the client has entered a variable, itself defined using a slider livelit, into the RGB
  splices. The A splice is filled directly with a slider livelit.
  The color livelit evaluates splices under its closure to display the
  computed color.
  The value of \li{default\_color} is determined by the livelit via a macro expansion step.}
  \label{fig:color}
\end{figure*}

We are not, of course, the first to seek to integrate direct manipulation user interfaces
into general-purpose programming systems.
Prior work on projectional editing
and active code completion, detailed in Sec. \ref{sec:related-work},
has also considered the problem of entering expressions
of certain types, like \li{Color},
using specialized GUIs integrated into a program editor.
The prior work most closely related to this paper is the {Graphite} system for Java \cite{Graphite},
demonstrated in Fig.~\ref{fig:color}(a).
Graphite allows library providers to associate GUIs, called \emph{palettes}, with types (using Java's class annotations).
Wherever an expression of a type equipped with a palette is needed,
i.e. wherever there is a \emph{hole} of that type in the program,
the programming environment offers the programmer the option, via the code completion menu,
to activate the palette.
Once the interaction is finished, the palette must generate a
Java expression to fill the hole.
Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a color palette invoked using Graphite.
When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 94)}\todo{correct code}{} is inserted at the cursor (not shown) and the palette disappears.

@@

Our focus in this paper is on addressing several non-trivial technical
deficiencies that limit both palette providers and palette clients in Graphite.
The system of \emph{live literals}, or \emph{livelits} introduced in
this paper and demonstrated in Fig.~\ref{fig:color}(b) improves upon Graphite
in each of the following ways.
\begin{enumerate}
  \setlength\itemsep{0.5em}
  \item \textbf{Decentralized Extensibility}:
    In Graphite, only one palette can be associated with a type, and this must
    be done at its definition site.
    By contrast, livelit definitions are orthogonal to type definitions.
    Clients invoke livelits explicitly via their names, which are prefixed by the \li{\$} character,
    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b).
    We call this \emph{decentralized extensibility}.
  \item \textbf{Persistence}: Graphite's palettes are {ephemeral},
  i.e. they disappear after the initial interaction,
  leaving behind only the generated textual code.
  Consequently, only the programmer that initially enters the expression
  benefits from the feedback and affordances that the palette provides.%
  \footnote{Graphite does include an \emph{ad hoc} mechanism that
  allows palettes to parse the code that is selected in the editor
  when the palette loads, but this requires that each palette implement
  a parser for the subset of Java used in the code that it generates,
  and therefore this mechanism is quite brittle. It is also difficult
  to persist GUI state that is not included in the generated code.}
  By contrast, livelits are persistent. We use a pure functional model-view-update-expand architecture
  (a variation on the Elm architecture \cite{ElmArchitecture}) as our GUI framework,
  so only the model needs to be persisted, rather than the full GUI state.
  We use the word ``literal'' rather than ``palette'' because, with persistence, livelits
  operate as graphical literal notation.

@@

  \item \textbf{Parameterization}: It is possible to define parameterized families of livelits.
  For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the lower and upper bound.
  Parameters behave like splices, differing in that livelit parameters can be partially applied in
  livelit abbreviations. For example, it is possible to define a livelit named \li{\$percentage}
  that operates as a \li{\$slider} from \li{0} to \li{100} by scoped abbreviation:
  \begin{lstlisting}[numbers=none]
  let livelit $percentage = $slider(0, 100) in ...
  \end{lstlisting}
  % Parameterization also underlies the hygiene mechanism, as we will discuss in Sec.~\ref{sec:livelit-definitions}.

  \item \textbf{Liveness}: Graphite's palettes do not have
  access to the run-time environment. By contrast, livelits can evaluate splices
  and therefore provide feedback related to the run-time behavior of the expression being generated.
  For example, in Fig.~\ref{fig:color}(b), the color that is displayed is determined by evaluating the RGBA
  component splices to numeric values.
  Evaluation occurs in a run-time environment (i.e. closure) determined by
  leaving the hole being filled by the livelit temporarily unfilled and then evaluating
  using a two-phased variation on the semantics for incomplete programs developed for Hazel \cite{HazelnutLive}.
  We will see more sophisticated examples later in the paper, including a livelit
  that appears inside a function that is called multiple times, leading to multiple closures that the client can
   select between.
\end{enumerate}

\noindent
\textbf{Outline.} The remainder of this paper is organized as follows. We begin in Sec.~\ref{sec:case-studies} with case studies
involving several non-trivial livelits, focusing on the livelit client's perspective.
In Sec.~\ref{sec:livelit-definitions}, we consider the livelit provider's perspective by detailing the livelit
definition mechanism by example.
In Sec.~\ref{sec:livelit-calculus}, we provide a formal account of the livelit mechanism as a typed lambda calculus.
This calculus formally specifies the essential aspects of livelit definition, representation, expansion,
and run-time closure collection independent of the particularities of GUI framework and other implementation details.
This calculus and the key metatheoretic properties has been mechanized in Agda.
In Sec.~\ref{sec:extensions}, we describe some extensions of the core calculus\todo{what are the extensions?}{}.
In Sec.~\ref{sec:implementation}, we provide a more detailed account of our implementation of livelits within Hazel,
and discuss several implementation details that would need to be considered by other implementations.
In Sec.~\ref{sec:related-work}, we compare livelits to related work.
Finally, we conclude in Sec.~\ref{sec:discussion} with a discussion of some present limitations and future work.

>>>>

\subsection{Background}
\begin{figure*}
  \begin{center}
    \begin{subfigure}[t]{0.5\textwidth}
      \centering
      \vspace{-3.4cm}
      \includegraphics[width=15pc]{graphite-color-palette.png}
    \end{subfigure}\begin{subfigure}[t]{0.5\textwidth}
      \centering
      \includegraphics[width=15pc]{slider-color-livelits.png}
    \end{subfigure}
  \end{center}
  \caption{
  (a) This figure, adapted from the prior work of \citet{Graphite},
  shows a Graphite palette associated with the \texttt{Color} class.
  The client can specify the RGBA components only using literal numbers.
  (b) A similar Hazel livelit for colors uses live splices for the RGBA components,
  so the client can enter any Hazel expression.
  In this case, the client has entered a variable, itself defined using a slider livelit, into the RGB
  splices. The A splice is filled directly with a slider livelit.
  The color livelit evaluates splices under its closure to display the
  computed color.
  The value of \li{default\_color} is determined by the livelit via a macro expansion step.}
  \label{fig:color}
\end{figure*}

We are not, of course, the first to seek to integrate direct manipulation user interfaces
into general-purpose programming systems.
Prior work on projectional editing
and active code completion, detailed in Sec. \ref{sec:related-work},
has also considered the problem of entering expressions
of certain types, like \li{Color},
using specialized GUIs integrated into a program editor.
The prior work most closely related to this paper is the {Graphite} system for Java \cite{Graphite},
demonstrated in Fig.~\ref{fig:color}(a).
Graphite allows library providers to associate GUIs, called \emph{palettes}, with types (using Java's class annotations).
Wherever an expression of a type equipped with a palette is needed,
i.e. wherever there is a \emph{hole} of that type in the program,
the programming environment offers the programmer the option, via the code completion menu,
to activate the palette.
Once the interaction is finished, the palette must generate a
Java expression to fill the hole.
Figure~\ref{fig:color}(a)\todo{use subfigure?}{}, adapted from this prior work, demonstrates a color palette invoked using Graphite.
When the user presses the \li{Enter} key, the Java expression \li{new Color(173, 173, 173, 85)} is inserted at the cursor (not shown) and the palette disappears.

@@

Our focus in this paper is on addressing several non-trivial technical
deficiencies that limit both palette providers and palette clients in Graphite
(and other projectional editing systems).
The system of \emph{live literals}, or \emph{livelits} introduced in
this paper and demonstrated in Fig.~\ref{fig:color}(b) improves upon Graphite
in each of the following ways.
\begin{enumerate}
  \setlength\itemsep{0.5em}
  \item \textbf{Decentralized Extensibility}:
    In Graphite, only one palette can be associated with a type, and this must
    be done at its definition site.
    By contrast, livelit definitions are orthogonal to type definitions.
    Clients invoke livelits explicitly via their names, which are prefixed by the \li{\$} character,
    e.g. \li{\$color} and \li{\$slider} in Fig.~\ref{fig:color}(b).
    We call this \emph{decentralized extensibility}.
  \item \textbf{Persistence}: Graphite's palettes are {ephemeral},
  i.e. they disappear after the initial interaction,
  leaving behind only the generated textual code.
  Consequently, only the programmer that initially enters the expression
  benefits from the feedback and affordances that the palette provides.%
  \footnote{Graphite does include an \emph{ad hoc} mechanism that
  allows palettes to parse the code that is selected in the editor
  when the palette loads, but this requires that each palette implement
  a parser for the subset of Java used in the code that it generates,
  and therefore this mechanism is quite brittle. It is also difficult
  to persist GUI state that is not included in the generated code.}
  By contrast, livelits are persistent. We use a pure functional model-view-update-expand architecture
  (a variation on the Elm architecture \cite{ElmArchitecture}) as our GUI framework,
  so only the model needs to be persisted, rather than the full GUI state.
  We use the word ``literal'' rather than ``palette'' because, with persistence, livelits
  operate as graphical literal notation.

@@

  \item \textbf{Parameterization}: It is possible to define parameterized families of livelits.
  For example, \li{\$slider} in Fig.~\ref{fig:color}(b) is parameterized by the lower and upper bound.
  Parameters behave like splices, differing in that livelit parameters can be partially applied in
  livelit abbreviations. For example, it is possible to define a livelit named \li{\$percentage}
  that operates as a \li{\$slider} from \li{0} to \li{100} by scoped abbreviation:
  \begin{lstlisting}[numbers=none]
  let $percentage = $slider 0 100 in ...
  \end{lstlisting}
  % Parameterization also underlies the hygiene mechanism, as we will discuss in Sec.~\ref{sec:livelit-definitions}.

  \item \textbf{Liveness}: Graphite's palettes do not have
  access to the run-time environment. By contrast, livelits can evaluate splices
  and therefore provide feedback related to the run-time behavior of the expression being generated.
  For example, in Fig.~\ref{fig:color}(b), the color that is displayed is determined by evaluating the RGBA
  component splices to numeric values.
  Evaluation occurs in a run-time environment (i.e. closure) determined by
  leaving the hole being filled by the livelit temporarily unfilled and then evaluating
  using a two-phased variation on the semantics for incomplete programs developed for the \Hazel programming environment \cite{HazelnutLive}.
  We will see more sophisticated examples later on, including a livelit
  that appears inside a function that is called multiple times, leading to multiple closures that the client can
   select between.
\end{enumerate}

\noindent
\textbf{Outline.} The remainder of this paper is organized as follows. We begin in Sec.~\ref{sec:case-studies} with case studies
involving several non-trivial livelits, focusing on the livelit client's perspective.
In Sec.~\ref{sec:livelit-definitions}, we consider the livelit provider's perspective by detailing the livelit
definition mechanism by example.
In Sec.~\ref{sec:livelit-calculus}, we provide a formal account of the livelit mechanism as a typed lambda calculus.
This calculus formally specifies the essential aspects of livelit definition, representation, expansion,
and run-time closure collection independent of the particularities of GUI framework and other implementation details.
This calculus and the key metatheoretic properties has been mechanized in Agda.
In Sec.~\ref{sec:implementation}, we provide a more detailed account of our implementation of livelits within Hazel,
and discuss several implementation details that would need to be considered by other implementations.
In Sec.~\ref{sec:related-work}, we compare livelits to related work.
Finally, we conclude in Sec.~\ref{sec:discussion} with a discussion of some present limitations and future work.
